Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-02-09 doc/invt.tex
vertical layout for figures

Generally, for an arbitrary updating matrix $\mathbf w$,
direct optimization of Eq. \eqref{eq:error_asym1}
often leads to a complicated expression for $\alpha(t)$.
%
In practice, it is also useful to find
a reasonably optimal $\alpha(t)$ with a simpler expression.
%
From Eq. \eqref{eq:alpha_opt},
we consider the following invert-time formula
%
\begin{equation}
\alpha(\tau) = \frac{1}{\lambda \, (\tau + t_0) },
\label{eq:alpha_invtlambda}
\end{equation}
%
for some value of $\lambda$.
%
Below we shall assume this functional form
and find the optimal value
in some cases.
%
We shall show that the optimal $\lambda$
generally depends on not only the spectrum
of eigenvalues and the sampling method,
but also other parameters,
such as the simulation length, $t$.
%
Note, however,
since Eq. \eqref{eq:alpha_invtlambda}
does not fix the endpoint value,
%
\begin{equation}
q(t) = \frac{ 1 } { \lambda }
\log\left(
  \frac{ t + t_0 } { t_0 }
\right),
\label{eq:qt_invtlambda}
\end{equation}
%
we have to include the residual error
during the optimization.

@@

We now turn to the nearest-neighbor updating scheme.
%
We again find good agreement
between simulation and analytical results,
although there are subtle differences.
%
A new feature is that the optimal $\lambda$
depends on the simulation length, $t$.
%
As shown in Fig. \ref{fig:err_nnbr},
the normalized error, $(t + t_0) \, \Err$,
hence the optimal value of $1/\lambda$,
depended non-trivially on the simulation length
in the global sampling case.
%
The optimal $1/\lambda$ gradually increased
with simulation length, which is in agreement with
the discussion in Sec. \ref{sec:optlambda}.
%
For the local sampling case, however,
the normalized error roughly maintained the sample shape,
and the optimal $\lambda$ is around $1.0$,
as suggested by Eq. \eqref{eq:lambda_nn_onestep}.

@@

In the above, we have assumed the inverse-time schedule,
Eq. \eqref{eq:alpha_invtlambda}.
%
In Fig. \ref{fig:alpha_cmp},
we compared the inverse-time schedule
with the optimal $\lambda$,
and the optimal schedule from integrating
Eq. \eqref{eq:q_opt}.
%
The difference was larger for the global sampling scheme.
%
\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/alphacmp.pdf}
  }
  \caption{
    \label{fig:alpha_cmp}
    Optimal schedules, $alpha(t)$, from Eq. \eqref{eq:q_opt},
    versus the inverse-time schedule,
    Eq. \eqref{eq:alpha_invtlambda},
    with the optimal parameter, $\lambda$.
    %
    For fair comparison,
    the $q(t) = \int_0^t \alpha(\tau) \, d\tau$
    were set to be the same
    for both schedules.
  }
\end{center}
\end{figure}

>>>>

Direct optimization of Eq. \eqref{eq:error_asym1}
for an arbitrary updating matrix $\mathbf w$
often leads to a complicated expression for $\alpha(t)$.
%
In practice, it is also useful to find
a reasonably optimal $\alpha(t)$ with a simpler expression.
%
From Eq. \eqref{eq:alpha_opt},
we consider the following invert-time formula
%
\begin{equation}
\alpha(\tau) = \frac{1}{\lambda \, (\tau + t_0) },
\label{eq:alpha_invtlambda}
\end{equation}
%
for some value of $\lambda$.
%
Below we shall assume this functional form
and find the optimal value
in some cases.
%
We shall show that the optimal $\lambda$
generally depends on not only the spectrum
of eigenvalues and the sampling method,
but also other parameters,
such as the simulation length, $t$.
%
Note, however,
since Eq. \eqref{eq:alpha_invtlambda}
does not fix the endpoint value,
%
\begin{equation}
q(t) = \frac{ 1 } { \lambda }
\log\left(
  \frac{ t + t_0 } { t_0 }
\right),
\label{eq:qt_invtlambda}
\end{equation}
%
we have to include the residual error
during the optimization.

@@

We now turn to the nearest-neighbor updating scheme.
%
We again find good agreement
between simulation and analytical results.
%
A new feature is that the optimal $\lambda$
depends on the simulation length, $t$.
%
As shown in Fig. \ref{fig:err_nnbr},
the normalized error, $(t + t_0) \, \Err$,
hence the optimal value of $1/\lambda$,
depended non-trivially on the simulation length
in the global sampling case.
%
The optimal $1/\lambda$ gradually increased
with simulation length, which is in agreement with
the discussion in Sec. \ref{sec:optlambda}.
%
For the local sampling case, however,
the normalized error roughly maintained the sample shape,
and the optimal $\lambda$ is around $1.0$,
as suggested by Eq. \eqref{eq:lambda_nn_onestep}.

@@

In the above, we have assumed the inverse-time schedule,
Eq. \eqref{eq:alpha_invtlambda}.
%
In Fig. \ref{fig:alpha_cmp},
we compared the inverse-time schedule
with the optimal $\lambda$,
and the optimal schedule from integrating
Eq. \eqref{eq:q_opt}.
%
The difference was larger for the global sampling scheme.
%
\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/alphacmp.pdf}
  }
  \caption{
    \label{fig:alpha_cmp}
    Optimal schedules, $\alpha(t)$, from Eq. \eqref{eq:q_opt},
    versus the inverse-time schedule,
    Eq. \eqref{eq:alpha_invtlambda},
    with the optimal parameter, $\lambda$.
    %
    For fair comparison,
    the $q(t) = \int_0^t \alpha(\tau) \, d\tau$
    were set to be the same
    for both schedules.
  }
\end{center}
\end{figure}
